Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-16 doc/invt.tex
invt.tex: introduction

\begin{abstract}
  Wang-Landau (WL) and metadynamics algorithms
  are widely-used techniques in free energy simulations.
  %
  These techniques allow one to sample a flat distribution
  along a quantity of interest
  by adaptively constructing a bias potential,
  as the negative potential of mean force (PMF).
  %
  However, the asymptotic error of the resulting PMF
  is determined by the magnitude of updating the bias potential,
  and in what fashion the magnitude is reduced
  over the simulation time.
  %
  In the Wang-Landau case,
  a simple and effective formula for the magnitude
  is given by the inverse of the number of simulation steps.
  %
  Here, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case,
  in which the bias potential at several neighboring bins
  are updated according to a predetermined window function,
  upon each visit.
  %
  Our analysis also suggests that in terms of convergence
  the WL algorithm represents one of the best of the class
  updating schemes among more sophisticated candidates.
\end{abstract}

@@

\section{Background}

\subsection{Flat-distribution sampling}

Given a system,
consider the problem of computing
the distribution, $p^*_i$
along a discrete quantity $i$.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy $E$
in lattice spin models; or the temperature index
in a simulated tempering simulation.
%
For the quantity of interest, $z$, is continuous,
which is typical in molecular systems,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin, $(z, z + dz)$.
%
In both cases,
the distribution can be normalized as
$\sum_{i = 1}^n p^*_i = 1$.

@@

To do so, we need to introduce a bias potential $V_i$,
such that the target distribution is changed to
%
\begin{equation}
  \pi_i \propto p^*_i \, e^{-V_i}.
  \label{eq:pi_p_phi1}
\end{equation}
%
Upon normalization $\sum_{i = 1}^n \pi_i = 1$,
we get
%
\begin{equation}
  \pi_i =
  \frac{ p^*_i \, e^{-V_i} }
  { \sum_{j = 1}^n p^*_j \, e^{-V_j} }.
  \label{eq:pi_p_phi}
\end{equation}
%
Particularly,
to achieve a flat distribution $\pi_i$,
the bias potential $V_i$
must coincide with $\log p^*_i$,
up to an additive constant.

Generally, we may wish to sample
a nearly flat distribution, $p_i$,
by adjusting the bias potential $V_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$V_i$ according to
$$
V_i^{\mathrm{new}}
=
V_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=V_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $-\log p^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process,
adopted by early multicanonical or entropic sampling,
can be inconvenient in practice,
for we need to make the period
as long as a multiple of
the autocorrelation time,
which is difficult to estimate for an unknown system.

\subsection{Wang-Landau algorithm}

The Wang-Landau (WL) algorithm is a convenient technique
of obtaining the desired bias potential $V_i$
on the fly.
%
To simplify the notation, we introduce
a shifted bias potential
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
%
and in $v_i$, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$).

In the WL algorithm, $v_i$ are updated
in each MC step $t$,
and $v_i(t)$ as a function of $t$
is updated as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
Note that the updating scheme only applies
to the $v_i$ at the current bin.
%The values of $v_j(t)$ at $j \ne i(t)$
%are kept unchanged.
%
One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
the trajectory is identical to $p_i$.
%
However, the updating scheme
makes $v_i(t)$ a fluctuating quantity,
and Eq. \eqref{eq:pi_p_v} no longer holds
at all times.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to reduce the error of $v_i$,
in order to get the unbiased distribution
$\ln p_i^*$ via Eq. \eqref{eq:v_def}.

@@

\subsection{$1/t$ formula}

A more effective way
of updating the $\alpha(t)$
is to discard the stage-wise setup,
and to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$ is the number of steps
from the beginning of the MC simulation,
which shall be referred to as the ``time'' below.
%
This surprisingly simple formula has attracted
several studies, notably the one by Zhou.
%
There are, however, some reservations about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a constant $C$
different from $1.0$.

Here we shall derive the optimal $\alpha(t)$
for the WL algorithm, Eq. \eqref{eq:wl_update}.
%
Further, we shall generalize the result
to a class of multiple-bin updating schemes,
in which an update affects
not only the current bin $i(t)$,
but also a few neighboring ones.
%(such schemes are designed
%to maintain a smooth bias potential).
%
We shall show that
the constant of proportionality, $1$,
in Eq. \eqref{eq:alpha_invt}
is optimal for the single-bin updating scheme,
Eq. \eqref{eq:wl_update}
(as in the WL algorithm),
but not necessarily so
for a multiple-bin updating scheme.
%
Below, we discuss the single-bin updating scheme
in Section \ref{sec:single-bin},
and the multiple-bin case
in Section \ref{sec:multiple-bin}.

\section{\label{sec:single-bin}
Single-bin updating scheme}

In this section,
we shall derive the optimal $\alpha(t)$
for the single-bin updating scheme,
Eq. \eqref{eq:wl_update}.
%
To do so,
we shall first express the error of $v(t)$
as a functional of $\alpha(t)$,
and then find the optimal value
by functional variation.

\subsection{Differential equation}

@@

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i
  =
  \frac{ p_i \, e^{-v_i} }
  { \sum_{j = 1}^n p_j \, e^{-v_j} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white.
and the correlation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.
%
More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$

\subsection{Linear approximation}

@@

\subsection{One-variable problem}

@@

\subsection{Optimal $\alpha(t)$ for the single-bin updating scheme}

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin updating scheme,
we find from Eq. \eqref{eq:dxdt_WL}
that with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
Thus, Eq. \eqref{eq:alpha_invt} is optimal
up to a constant $t_0$. %defining the origin of $t$.

\section{\label{sec:multiple-bin}
Multiple-bin update}

We now consider the multiple-bin updating scheme.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential there,
but also those for a few neighboring bins $i$.
%
The update magnitudes are given by a matrix $\mathbf w$
with elements $w_{ij}$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t) =
  \sum_{j=1}^n \alpha(t) \, h_j(t) \frac{ w_{ij} } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

@@

\subsection{Updating matrix $\mathbf w$}

@@

\subsection{Linearization and decoupling}

@@

\subsection{Minimization of the error function}

@@

\subsection{White-noise approximation}

In the asymptotically regime,
where $t \gg 1/\max \alpha(t)$,
we can approximate $\kappa_i(t)$ as
%
\begin{equation}
  \kappa_i(t) \approx \Gamma_i \, \delta(t)
  \label{eq:kappa_delta}
\end{equation}
%
where, $\Gamma_i = \int_{-\infty}^\infty \kappa_i(t) \, dt$.
%
Then, Eq. \eqref{eq:error_mbin} is simplified as
%
\begin{align}
  \mathcal E
  &=
  \int_0^t
  \sum_{i = 2}^n
  \Gamma_i \, \dot u_i^2\bigl( q(t') \bigr) \, dt',
  \label{eq:error_mbin1}
\end{align}
%
%
\note{This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_i(\omega) = \Gamma_i
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_i(\omega) = \Gamma_i + \Gamma^{(2)}_i \omega^2 + \dots,
  $$
  which leads to a correction to $\mathcal E$
  $$
  \Gamma^{(2)}_i
  \int_0^t \ddot u_i^2\bigl( q(t') \bigr) \, dt'
  \propto
  \alpha^2 \, \mathcal E.
  $$
  This correction is of order $\alpha^2$
  compared to the main contribution in Eq. \eqref{eq:error_mbin1}.
  %
  Thus, we can assume Eq. \eqref{eq:kappa_delta},
  once $\alpha(t)$ has dropped below a certain level.
  \\
}
\note{
  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{i=1}^n
      \Gamma_i \, \lambda_i \,
      u_i\bigl( q(\tau) \bigr) \,
      \ddot u_i\bigl( q(\tau) \bigr) = 0.
    \label{eq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_i$ as the position,
  and $\Gamma_i \, \lambda_i$ as the mass,
  then the right-hand side give the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}
%
For example,
if we assume perfect sampling,
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  p_i \, (1 - p_i) \, \delta_{ij} \, \delta(t - t').
  \label{eq:zz_perfect}
\end{equation}
%
Thus, from Eqs. \eqref{eq:eig_orthonormal_rows} and \eqref{eq:eta_def},
we get
%
\begin{equation}
  \kappa_i(t - t')
  =
  \sum_{j,k = 1}^n
  \frac{ \phi_{ij} } { p_j }
  \frac{ \phi_{ik} } { p_k }
  \langle \zeta_j(t) \, \zeta_k(t') \rangle
  %
  \approx \delta(t - t').
  \label{eq:kappa_perfect}
\end{equation}
%
where we have assumed the large $n$ limit
such that $p_i \ll 1$
in the last expression.

@@

\subsection{\label{sec:optWL}
Optimality of the single-bin updating scheme}

>>>>

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques in free energy simulations.
  %
  They allow one to sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing a bias potential,
  which is the negative potential of mean force (PMF).
  %
  The two algorithms differ by
  the manner of updating the bias potential:
  %
  in the WL case, each simulation step triggers
  an update of the bias potential to only
  the bin containing the current $z$,
  while in the metadynamics case,
  the bias potential of several neighboring bins
  are affected as well.
  %
  The asymptotic error of the resulting PMF
  is determined by the updating magnitude,
  and the rate of the reduction of the magnitude
  over the simulation time.
  %
  In the WL case,
  a simple and effective formula for the magnitude
  is given by the inverse of the number of simulation steps.
  %
  In this study, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case.
  %
  Further, we compared the two algorithms, and found that
  although metadynamics produces a smoother PMF profile,
  it is less effective in reducing errors with the shorter wavelengths.
  %
  Thus, it relies more heavily on the assumption of
  the smoothness of the PMF profile,
  especially with a wide updating window.
\end{abstract}

@@

\section{Introduction}

Free energy calculation\cite{frenkel} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute a distribution, $p^*(z)$,
along a quantity of interest $z$, and to see how
the distribution changes with the external conditions,
such as temperatures and pressures.
%
The negative logarithm $-\log p^*(z)$,
or the potential of mean force (PMF),
defines a free energy.
%
A straightforward approach of computing the PMF
is to run multiple independent simulations
under different conditions,
and then piece together the information
gathered there.
%
This solution is often unsatisfactory
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially alter
the target distribution,
such that we can, in a single simulation, sample
a flat distribution\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002}
along $z$ over a wide range.
%
This alleviates the above problem of local trapping
as the system is now forced to travel along the $z$ direction,
which often represents a slow reaction coordinate.
%
It, however, requires one to efficiently construct a
bias potential for an often unknown system.
%
Note that the bias potential is directly related to the PMF
as the former must precisely offset the latter
to achieve a flat distribution.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{laio2002}
are two widely-used techniques for this purpose,
in which a bias potential is actively build-up
based on the simulation history.
%
In essence, these techniques
encourage the system to leave a region that has been visited
by elevating the bias potential there.
%
To be effective, the updates to bias potential
are made at every few steps,
while a side effect is that the underlying dynamics
is no longer a equilibrium sampling
required by statistical mechanics.
%
Thus, one has to reduce the magnitude of updating
over the course of simulation.

Naturally, the manner of reducing
the updating magnitude\cite{
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}
determines the precision of the final bias potential,
hence that of PMF.
%
For the WL algorithm, it is found
that asymptotically the optimal magnitude of updating $\ln f$
should be given by the
inverse time\cite{
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008},
where ``time'' is defined as
the number of simulation steps
divided by the number of bins.

In this study,
we shall give a proof of the inverse-time formula,
and try to generalize it to
updating schemes involving neighboring bins,
such as those used in metadynamics.
%
As a byproduct, we shall show that
the single-bin scheme used by the WL algorithm
is generally one of the best
updating schemes in terms of convergence.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

\section{\label{sec:theory}
Theory}

In this section,
we first review the basics and fix notations
in Sec. \ref{sec:background}.
%
We then prove the optimality
of the inverse-time formula
for the single-bin WL algorithm
in Sec. \ref{sec:single-bin}.
%
Next, we discuss the generalized
multiple-bin schemes
in Secs. \ref{sec:multiple-bin}
and \ref{sec:band-matrix}.
%

\subsection{\label{sec:background}
Background}

\subsubsection{Flat-distribution sampling}

Given a system,
consider the problem of computing
the distribution, $p^*_i$
along a discrete quantity $i$.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy $E$
in lattice spin models; or the temperature index
in a simulated tempering simulation.
%
For the quantity of interest, $z$, is continuous,
which is typical in molecular systems,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin, $(z, z + dz)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

@@

To do so, we need to introduce a bias potential $V_i$
to modify the target distribution to
%
\begin{equation}
  \pi_i \propto p^*_i \, e^{-V_i}.
  \label{eq:pi_p_phi1}
\end{equation}
%
Upon normalization, $\sum_{i = 1}^n \pi_i = 1$,
we get
%
\begin{equation}
  \pi_i =
  \frac{ p^*_i \, e^{-V_i} }
  { \sum_{j = 1}^n p^*_j \, e^{-V_j} }.
  \label{eq:pi_p_phi}
\end{equation}
%
Particularly,
to achieve a flat distribution $\pi_i$,
the bias potential $V_i$
must coincide with $\log p^*_i$,
up to an additive constant.

Generally, we may wish to sample
a nearly flat distribution\cite{dayal2004, trebst2004}, $p_i$,
by adjusting the bias potential $V_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$V_i$ according to
$$
V_i^{\mathrm{new}}
=
V_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=V_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $-\log p^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process,
adopted by early multicanonical or
entropic sampling\cite{berg1992, lee1993},
can be inconvenient in practice,
for we need to make the period
as long as a multiple of
the autocorrelation time,
which is difficult to estimate for an unknown system.

For later convenience, we introduce
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
are constant during the course of simulation,
updates to $V_i$ are equivalent to those to $v_i$.
%
In terms of $v_i$, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$).

\subsubsection{Updating schemes}

In the WL algorithm\cite{wang2001, wang2001pre},
$v_i$ are updated
in each MC step $t$,
and $v_i(t)$ as a function of $t$
is updated as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to the above scheme as the single-bin scheme,
for it applies only to the bias potential
at the current bin, $i(t)$,
%
By contrast, a multiple-bin scheme
means the general case in which
several neighboring bins are updated.

%The values of $v_j(t)$ at $j \ne i(t)$
%are kept unchanged.

One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
the trajectory is identical to $p_i$.
%
However, the continuous update
makes $v_i(t)$ a fluctuating quantity,
and Eq. \eqref{eq:pi_p_v} no longer holds
at all times.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to reduce the error of $v_i$,
in order to get the unbiased distribution
$\ln p_i^*$ via Eq. \eqref{eq:v_def}.

@@

\subsubsection{$1/t$ formula}

A more effective way
of updating the $\alpha(t)$
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$ is the number of steps
from the beginning of the MC simulation,
which shall be referred to as the ``time'' below.
%
This surprisingly simple formula has attracted
several studies, notably the one by Zhou.
%
There are, however, some reservations about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a constant $C$
different from $1.0$.

%Here we shall derive the optimal $\alpha(t)$
%For the WL algorithm, Eq. \eqref{eq:wl_update}.
%%
%Further, we shall generalize the result
%To a class of multiple-bin schemes,
%In which an update affects
%Not only the current bin $i(t)$,
%But also a few neighboring ones.
%%(such schemes are designed
%%to maintain a smooth bias potential).
%%
%We shall show that
%The constant of proportionality, $1$,
%In Eq. \eqref{eq:alpha_invt}
%Is optimal for the single-bin scheme,
%Eq. \eqref{eq:wl_update}
%(as in the WL algorithm),
%But not necessarily so
%For a multiple-bin scheme.
%%
%Below, we discuss the single-bin scheme
%In Section \ref{sec:single-bin},
%And the multiple-bin case
%In Section \ref{sec:multiple-bin}.

\subsection{\label{sec:single-bin}
Single-bin scheme}

In this section,
we shall derive the optimal $\alpha(t)$
for the single-bin scheme,
Eq. \eqref{eq:wl_update}.
%
To do so,
we shall first express the error of $v(t)$
as a functional of $\alpha(t)$,
and then find the optimal value
by functional variation.

\subsubsection{Differential equation}

@@

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i
  =
  \frac{ p_i \, e^{-v_i} }
  { \sum_{j = 1}^n p_j \, e^{-v_j} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the correlation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.
%
More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$

\subsubsection{Linear approximation}

@@

\subsubsection{One-variable problem}

@@

\subsubsection{Optimal $\alpha(t)$ for the single-bin scheme}

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin scheme,
we find from Eq. \eqref{eq:dxdt_WL}
that with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
Thus, Eq. \eqref{eq:alpha_invt} is optimal
up to a constant $t_0$. %defining the origin of $t$.

\subsection{\label{sec:multiple-bin}
Multiple-bin scheme}

We now consider the multiple-bin scheme.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential there,
but also those for a few neighboring bins $i$.
%
The update magnitudes are given by a matrix $\mathbf w$
with elements $w_{ij}$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t) =
  \sum_{j=1}^n \alpha(t) \, h_j(t) \frac{ w_{ij} } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

@@

\subsubsection{Updating matrix}

@@

\subsubsection{Linearization and decoupling}

@@

\subsubsection{Minimization of the error function}

@@

\subsubsection{White-noise approximation}

In the asymptotic regime,
we can approximate $\kappa_i(t)$ as
%
\begin{equation}
  \kappa_i(t) \approx \Gamma_i \, \delta(t)
  \label{eq:kappa_delta}
\end{equation}
%
where, $\Gamma_i = \int_{-\infty}^\infty \kappa_i(t) \, dt$.
%
Then, Eq. \eqref{eq:error_mbin} is simplified as
%
\begin{align}
  \mathcal E
  &=
  \int_0^t
  \sum_{i = 2}^n
  \Gamma_i \, \dot u_i^2\bigl( q(t') \bigr) \, dt'.
  \label{eq:error_mbin1}
\end{align}
%
The next order correction to Eq. \eqref{eq:kappa_delta}
would lead to a correction to $\mathcal E$
that is about $\alpha_{\max}^2$ as large.
%
\note{This is equivalent to saying that the Fourier transform
  $$
  \tilde \kappa_i(\omega) = \Gamma_i
  $$
  is a roughly constant.
  %
  For the next order correction we would have
  $$
  \tilde \kappa_i(\omega) = \Gamma_i + \Gamma^{(2)}_i \omega^2 + \dots,
  $$
  which leads to a correction to $\mathcal E$
  $$
  \Gamma^{(2)}_i
  \int_0^t \ddot u_i^2\bigl( q(t') \bigr) \, dt'
  \propto
  \alpha^2 \, \mathcal E.
  $$
  This correction is of order $\alpha^2$
  compared to the main contribution in Eq. \eqref{eq:error_mbin1}.
  %
  Thus, we can assume Eq. \eqref{eq:kappa_delta},
  once $\alpha(t)$ has dropped below a certain level.
  \\
}
\note{
  The extremal property, Eq. \eqref{eq:optimal_mbin},
  now becomes
  %
  \begin{align}
    \sum_{i=1}^n
      \Gamma_i \, \lambda_i \,
      u_i\bigl( q(\tau) \bigr) \,
      \ddot u_i\bigl( q(\tau) \bigr) = 0.
    \label{eq:optimal_mbin1}
  \end{align}
  %
  If we interpret $u_i$ as the position,
  and $\Gamma_i \, \lambda_i$ as the mass,
  then the right-hand side give the virial.
  %
  Thus, the extremal condition demands
  the total virial to be zero.
  $$\,$$
}
%
For example,
if we assume perfect sampling,
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  p_i \, (1 - p_i) \, \delta_{ij} \, \delta(t - t').
  \label{eq:zz_perfect}
\end{equation}
%
Thus, from Eqs. \eqref{eq:eig_orthonormal_rows} and \eqref{eq:eta_def},
we get
%
\begin{equation}
  \kappa_i(t - t')
  =
  \sum_{j,k = 1}^n
  \frac{ \phi_{ij} } { p_j }
  \frac{ \phi_{ik} } { p_k }
  \langle \zeta_j(t) \, \zeta_k(t') \rangle
  %
  \approx \delta(t - t').
  \label{eq:kappa_perfect}
\end{equation}
%
where we have assumed the large $n$ limit
such that $p_i \ll 1$
in the last expression.

@@

\subsubsection{\label{sec:optWL}
Optimality of the single-bin scheme}

We now show
that the single-bin scheme
is an optimal scheme.
%
Consider a class of $\mathbf w$ matrices
with the same set of eigenvectors,
hence same $\Gamma_i$'s,
but with different eigenvalues, $\lambda_i$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of $\beta_i \ge 0$
%
\begin{align}
&
\left( \int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
                     \dot u_i^2\bigl( q(t') \bigr)
\right)
\left(
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \, c_i^2
\right)
\notag
\\
&
\ge
\left(
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \, c_i \,
                     \dot u_i\bigl( q(t') \bigr)
\right)^2
\notag
\\
&
=
\left(
  \sum_{i = 2}^n \Gamma_k \, c_i
    \left[
      1 - e^{ -\lambda_i q(t) }
    \right]
\right)^2.
\notag
\end{align}
%
\note{A short proof of the above version
  of the Cauchy-Schwarz inequality
  is the following.
  %
  First, we have
$$
\int_0^t
  dt' \sum_{i = 2}^n \Gamma_i \,
    \left( \dot u_i \, X - c_i \right)^2
\ge 0,
$$
  for any $X$.
  %
  Thus, the right-hand side,
  as a quadratic polynomial of $X$,
  must have a non-positive discriminant.
  %
  This condition gives the desired inequality.
  \\%
}
%
Note that the last expression depends on
$q(t')$ only through the end point value,
which is fixed in the variation process.
%
Thus, the inequality sets a lower bound
of the error $\mathcal E$
expressed in Eq. \eqref{eq:error_mbin1},
%
The equality is achieved
if $\dot u_i\left( q(t') \right) = c_i$
for all $i = 2, \dots, n$ at any $t'$,
up to a multiplicative factor.
%
Solving this set of equations yields
$$
\alpha(t') = \frac{              1              }
                  { \lambda_i \, (t' + t_{i0} ) },
\quad
\mathrm{with\;\;}
t_{i0} = \frac{             t            }
              { e^{ \lambda_i q(t) } - 1 }.
$$
Such a solution is possible only if
\begin{equation}
\lambda_2 = \dots = \lambda_n.
\end{equation}
%
\note{
  We add some steps for solution.
  Integrating $\dot u_i = c_i$ yields
  \begin{equation}
  u_i(t') = c_i \left(t' + t_{i0} \right).
  \label{eq:ui_solution}
  \end{equation}
  Given that $u_i(t) = 1$ and $u_i(0) = e^{-\lambda_i \, q(t)}$,
  we have
  $$
  c_i = \frac{ 1 }{ t + t_{i0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{i0} } { t + t_{i0} }
  =
  e^{ -\lambda_k \, q(t) }.
  $$
  Taking the logarithm of Eq. \eqref{eq:ui_solution} yields
  $$
  -\lambda_i \, q(t) + \lambda_i \, q(t')
  = \log c_i + \log\left( t' + t_{i0} \right).
  $$
  Differentiating this with respect to $t'$ yields
  $$
  \lambda_i \, \alpha(t') = \frac{ 1 } { t' + t_{i0} }.
  $$
}

Now an updating matrix $\mathbf w$ satisfying this condition
is essentially a multiple of the identity matrix
(since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_i$,
we can freely set it to $\lambda_2$
without changing the nature of the updating scheme).
%
Thus, in terms of convergence,
the single-bin scheme
(as adopted in the WL algorithm)
is generally the most efficient
updating scheme.

\subsubsection{Optimal coefficient of proportionality}
